Circularity in Voxel-Based Analysis

### 2011-10-17 ANTsTBSS/paperRevision.tex
ENH: Adding revision stuff

\newcommand{\lstsetcpp}
{
\lstset{frame = tb,
        framerule = 0.25pt,
        float,
        fontadjust,
        backgroundcolor={\color{listlightgray}},
        basicstyle = {\ttfamily\scriptsize},
        keywordstyle = {\ttfamily\color{listkeyword}\textbf},
        identifierstyle = {\ttfamily},
        commentstyle = {\ttfamily\color{listcomment}\textit},
        stringstyle = {\ttfamily},
        showstringspaces = false,
        showtabs = false,
        numbers = none,
        numbersep = 6pt,
        numberstyle={\ttfamily\color{listnumbers}},
        tabsize = 2,
        language=[ANSI]C++,
        floatplacement=!h,
        caption={},
        captionpos=b,
        }
}

@@

\title{Circularity in Population-Based FA Analysis}

@@

\begin{abstract}
Template-based normalization plays a critical role in many population
studies based on diffusion tensor imaging (DTI).  Several current
approaches compute the fractional anisotropy (FA) from the DTI and then
directly register the FA image to a FA template before performing
statistical comparison between different groups within a population.  In
this paper, we demonstrate that this analysis strategy induces selection
bias that varies in strength with the similarity metric used in
registration.  Specifically, we show that the popular sum of squared
difference intensity metric maximizes effect size, rather than
anatomical alignment, and consistently overestimates statistical
significance.  Consequently, we advocate a more cautious approach
that normalizes a population of FA images through the associated 
anatomical images.  We show that the Symmetric Group Normalization
strategy for unbiased template construction from anatomical (e.g.
T1-weighted) images provides a valid and unbiased coordinate system for
statistical inference of FA differences and yet retains detection power
in comparison to FA-based normalization methods.  We establish this
outcome by using open data comprised of normal controls from the International
Neuroimaging Data-sharing Initiative as well as a cohort of 16 traumatic
brain injury survivors compared with 17 control subjects.  Alignment
comparisons with the normalization component of the popular tract-based
spatial statistics (TBSS) show that our anatomically-based template
approach produces good alignments despite the absence of any direct
FA-to-FA registration.  Of practical consideration we note that the
software tools discussed in this work are provided to the public as part
of the open source Advanced Normalization Tools (ANTs) repository.  In conclusion, we recommend against circular analysis strategies that select image similarity metrics that are tied directly to the statistical functions used to assess population effects.

%We propose a modification of the popular tract-based spatial statistics (TBSS) framework \citep{Smith2006} for investigating diffusion tensor derived scalar measures across populations.   Whereas the current crucial alignment step is performed typically using the derived fractional anisotropy images, we propose using the Symmetric Group Normalization strategy for anatomical unbiased template construction from T1-weighted images which serves as the coordinate system for statistical inference.   A potential benefit of this modified approach is reduction of the general confounding effect of minimizing data differences via nonrigid registration which are potentially the precise differences to be statistically quantified.  Alignment performance is evaluated using open data sets from the International Neuroimaging Data-sharing Initiative.  This assessment demonstrated improved alignment in both fractional anisotropy (FA) and mean diffusivity (MD) images versus standard TBSS.  Further analyses evidencing increased regional sensitivity in peripheral white matter to FA differences include a DTI-based assessment of a cohort of 16 traumatic brain injury survivors compared with 17 control subjects.  Similar to the Oxford Centre for Functional MRI of the Brain (FMRIB) who provide TBSS to the public, the software tools discussed herein are provided to the public as part of the open source Advanced Normalization Tools (ANTs) repository. 
\end{abstract}

\begin{keyword}
Advanced Normalization Tools (ANTs) \sep diffusion tensor imaging \sep fractional anisotropy \sep template construction
%% keywords here, in the form: keyword \sep keyword
\end{keyword}

@@

\section{Introduction}
%State the objectives of the work and provide an adequate background, avoiding a detailed literature survey or a summary of the results.

The seminal work of Basser et al. \citep{Basser1994a,Basser1994} established diffusion tensor imaging (DTI) as a viable investigatory MRI technique.  DTI's sensitivity to brain architecture \citep{Basser1996,Assaf2008} enables promising analysis possibilities assessing neuro-structural differences in cross-population studies \citep{Kubicki2005,Arnone2008,Kantarci2010,Rametti2010} including subjects with traumatic brain injury (TBI) \citep{Caeyenberghs2010,Jiang2010,Warner2010}.  Two popular comprehensive software packages for assessing population FA differences include the SPM Matlab-based toolkit%
\footnote{
http://www.fil.ion.ucl.ac.uk/spm/
}
for voxel-based morphometry (VBM)
and the tract-based spatial statistics (TBSS) framework \citep{Smith2006}.%
\footnote{
http://www.fmrib.ox.ac.uk/fsl/tbss/index.html
} 
VBM \citep{Ashburner2001} continues to find application to FA studies \citep{Kakeda2010,Takao2010,Preziosa2011} despite concerns about its general validity \citep{Bookstein2001,Davatzikos2004} and with respect to DTI \citep{Jones2005,Chung2008}.  The TBSS framework was partially developed in response to these concerns in which voxel values are projected onto a template white matter skeleton for increased statistical power.  
%(e.g. \citep{Arnone2008,Bodini2009}).

Although there are substantive differences in normalization
between these and other analysis approaches, a core
commonality includes direct FA-to-FA registration {\color{red}{using various similarity metrics}}.  The sum of squared differences (SSD) image similarity metric is perhaps the easiest to interpret (it drives the image difference to zero), is computationally efficient (just compute the image difference and gradient) and is therefore widely used.  Some of the most popular registration methods (SPM, Demons, FSL's FNIRT) rely on this metric and yield reasonable performance levels \citep{Klein2009}.%
\footnote{
In addition to TBSS and SPM, other popular packages in the neuroimaging community 
which use the SSD metric include AIR \citep{Woods1998} and LDDMM \citep{Beg2005}.
}

However, suppose the SSD metric is used to find a set of $M$ transformations that map a population of $M$ FA images, $\{I_1, I_2, \ldots, I_M\}$, to a representative FA template, $J$, for statistical testing between groups. This results in the transformations,
$\boldsymbol{\mathcal{T}} = \{\mathcal{T}_1, \mathcal{T}_2,\ldots,\mathcal{T}_M \}$, which minimize the SSD metric over the population, i.e.
\begin{align}
  \boldsymbol{\mathcal{T}} = \underset{\mathcal{T}_1, \ldots, \mathcal{T}_M}{\operatorname{argmin}}
    \sum_{m=1}^M\sum_{n=1}^N \left( I^n_m( \mathcal{T}_m(\mathbf{x}) ) - J \right)^2.
\end{align}
Switching the order of the summations (since the transformations are mutually independent) and recognizing that a principal criterion for the selection of $J$ is such that it be a good approximation of the mean of the aligned images, it becomes apparent that the transformation solution
\begin{align}\label{eq:variance}
  \boldsymbol{\mathcal{T}} = \underset{\mathcal{T}_1, \ldots, \mathcal{T}_M}{\operatorname{argmin}}
    \sum_{n=1}^N \underbrace{\sum_{m=1}^M \left( I^n_m( \mathcal{T}_m(\mathbf{x}) ) - J \right)^2}_{
    \propto\mathrm{\,voxelwise\,variance}}
\end{align}
is that which minimizes the voxelwise variance.  Reducing variance will also directly affect the value of a standard population statistic, the Student's $t$-test.  Other standard similarity metrics will also exhibit this
trend to some degree as they minimize some local measure of intensity-based distance between the subject and the template.

Thus, instead of normalization based on anatomical alignment followed by
statistical testing, direct FA normalization explicitly conflates
anatomical alignment with optimizing the very statistical testing
results that one will ultimately use to assess hypotheses.  This immediately evokes a sense of circularity in the analysis \citep{Kriegeskorte2010}.
According to Kriegeskrote et al., ``An analysis is circular (or nonindependent) if it is based on data that were selected for showing the effect of interest or a related effect.''  As illustrated with
Eqn. \ref{eq:variance}, direct FA-to-FA registration will produce transformations that align (or select) voxels such that the statistical testing result (or the effect of interest) is increased (assuming one is using Student's 
$t$-test).  While the SSD metric is able to provide visually attractive alignments across subjects, the voxel-level details (that is, which voxels are matched where) are, as we show in the following simulation study, strikingly important in terms of the effect on detection power.

%This results in a significant increase in the number of false positives for this type
%of testing which we illustrate with the following example.

\paragraph{Simulated Data Example}

\begin{figure}
\begin{center}
\begin{tabular}{c}
  \includegraphics[width=85mm]{simulatedData.pdf}
\end{tabular}
\caption{(a) Axial slice from the discrete BrainWeb data set.  From (a) 
we create the (b) simulated template and sample (c) control and (d) patient images with both 
additive noise and subsequent Gaussian smoothing. }
\label{fig:simulated_data}
\end{center}
\end{figure}

\begin{figure}
\begin{center}
\begin{tabular}{c}
  \includegraphics[width=85mm]{simulatedDataWarped.pdf}
\end{tabular}
\caption{Warping the image in Fig. \ref{fig:simulated_data}(c) to the template in 
  Fig. \ref{fig:simulated_data}(b) using the SSD similarity metric produces the image shown in
  (a).  The corresponding deformation is given in (b).  Although the images were anatomically
  aligned prior to invoking ANTs registration, deformation is driven by the similarity metric
  causing the accumulation of Type 1 errors.}
\label{fig:simulated_data_warped}
\end{center}
\end{figure}

Axial slice number 159 was extracted from \verb#subject_04# of the discrete BrainWeb database
\citep{Aubert-Broche2006} which is shown in Fig. \ref{fig:simulated_data}.  After thresholding out 
the white matter and assigning all white matter voxels a value of 0.7, 10\% of these white 
matter voxels were randomly selected to be assigned a value of 0.6 in the simulated control 
template and a value of 0.3 in the simulated patient template.  
Note that these values were arbitrarily selected to be more or less in the range of typical
FA values.  Gaussian noise (mean = 0.0, standard deviation = 0.1) was added to the white
matter template followed by a Gaussian smoothing (standard deviation = 1.5 mm) shown in Fig. \ref{fig:simulated_data}(b).  A simulated control cohort was created consisting of 20 subjects.  
Starting with the simulated control template, each of the 20 subjects is created by adding
Gaussian noise (mean = 0.0, standard deviation = 0.1) followed by 
Gaussian smoothing (standard deviation = 1.5 mm).  Thus, each subject is differentiated only by a unique random draw from the noise distribution and there is no real shape difference between subjects.  A simulated patient cohort of 20 subjects
was created similarly using the simulated patient template.  A sample control subject and sample
patient subject are given, respectively, in Fig. \ref{fig:simulated_data}(c) and (d).

\begin{figure*}
\begin{center}
\begin{tabular}{c}
%  \includegraphics[width=180mm]{simulatedTesting.pdf}
  \includegraphics[width=130mm]{simulatedTesting.pdf}
\end{tabular}
\caption{Student's $t$-test and the nonparametric Wilcoxon rank-sum test results (one-tailed, $\alpha = 0.95$) for the simulated experiment adjusted for multiple comparisons (FDR).
The true significance curves are drawn in black.  Elevated statistical power is seen with all 4 
         metrics for both tests although the effect is less in the more robust Wilcoxon results. Not shown are the significance curves for the Brunner-Munzel testing \citep{Rorden2007} which showed similar effects to the results on the left.
        }
\label{fig:simulated_testing}
\end{center}        
\end{figure*}

After creation of the simulated data set, we performed voxelwise testing (both $t$-testing
and Wilcoxon rank-sum testing, aka the Mann-Whitney test, one-tailed,
$\alpha = 0.95$) with false discovery rate correction within the white matter to get ``true'' cumulative distribution functions (CDFs).  The CDF vertical axis shows the cumulative number of voxels with corrected $p$-values at or below a given significance level (horizontal axis) and thus may be used as a surrogate measure for spatial sensitivity.  We denote these graphs as "statistical significance curves."  
The true curves for the simulated experiment are plotted in black in Fig. \ref{fig:simulated_testing}.  
We note that the amount of smoothing increases the proportion of significant voxels beyond
the prescribed 10\%.

To illustrate the effect of similarity metric on statistical significance, we performed ANTs registration \citep{Avants2011} of each subject to the template
(Fig. \ref{fig:simulated_data}(b)) using cross correlation (CC), mutual information (MI), 
Demons \citep{Thirion1998}, and sum of squared differences (SSD). A representative deformed image
using the SSD metric and its corresponding deformed grid is given in 
Fig. \ref{fig:simulated_data_warped}. 
The resulting statistical significance curves are plotted with the true significance curves in Fig. 
\ref{fig:simulated_testing}.

This example demonstrates the undesirable phenomenon previously
described.  Even though these images were anatomically aligned prior
to registration, the resulting deformations (falsely) increased
detection power.  This effect can be explained in the context of
Eqn. (\ref{eq:variance}) and by the fact that we added noise to the
underlying distribution which allows the deformable mapping algorithms
the freedom to transform the positions of the most "powerful" voxels
to minimize variance across the population (and increase the $t$-test
statistics).  Although we do not show the results, additional
experiments show that decreasing the standard deviation of the noise
causes an increase in separation between the SSD/Demon's metrics and
the MI and CC metrics.  MI and CC are closer to the the true
significance curve.  Alternatively, if we increase the noise level,
the same ranking occurs with even more spread between significance curves.

Common similarity metric functions can therefore cause false positives by minimizing local intensity 
variance between the subjects and template which explicitly maximizes statistical
testing results in contrast to maximizing anatomical alignment.  This effect is strongest in 
the SSD and Demons metrics and weakest in the MI and CC metrics likely due
to the degree of correspondence to Eqn. (\ref{eq:variance}).   This is also supported by contrasting
the $t$-test results given in Fig. \ref{fig:simulated_testing}(a) with the non-parametric Wilcoxon
rank-sum testing (Fig. \ref{fig:simulated_testing}(b)).  The rank test does not explicitly use variance in its assessment and thus exhibits far less statistical exaggeration even though it offers roughly 95\%
the power of the $t$-test assuming certain assumptions are met \citep{Siegel1988,Rorden2007}.

Although various processing choices mitigate these effects, such as Gaussian smoothing, nonparametric testing, and the use of more robust similarity metrics, such choices are often made on an ad hoc basis.  
In contrast, in this work we strongly advocate the general application
%A more principled approach is the use of an anatomical, i.e. T1-weighted, unbiased shape/intensity 
%template constructed from the population images.  DTI-derived scalar images can then be mapped 
%to their corresponding T1-weighted image in a highly constrained manner and subsequently 
%mapped to the template via nonrigid registration of the T1-weighted image.  This minimizes 
%possible complications associated with direct FA-to-FA image registration.
%We recognize that using the anatomical images in the alignment process has been previously proposed (in fact, such a suggestion occurs in the {\it Future directions} section of the original TBSS paper).  
of a high performance \citep{Klein2009}, nonrigid registration algorithm \citep{Avants2011}, which forms the underlying machinery for an optimal template construction algorithm \citep{Avants2010}, in facilitating alignment between DTI-derived scalar images for cross-population studies.  The key point is that the FA images are mapped to the common reference space via intra-subject DT-to-T1 mapping followed by T1-to-T1 alignment.  The T1-to-T1 step is (statistically) independent of the FA values which are the ultimate statistical target of hypothesis testing in FA population studies.  Below, we show that this same pattern of effects is reproduced in real data and detail our proposed solution.

\begin{figure*}
\begin{center}
\begin{tabular}{c}
%  \includegraphics[width=170mm]{workflow.pdf}
  \includegraphics[width=130mm]{workflow.pdf}
\end{tabular}
\caption{Possible ANTs-based TBSS workflow using SyGN.  In contrast to standard TBSS in which an individual subject's FA is registered to the FMRIB58 template
(transform represented with the blue arrow), our proposed modification first constructs an ANTs optimal anatomical template as illustrated in the center of the figure.  Each FA image is then rigidly mapped, with distortion correction, to the individual T1 using the white matter mask as shown on the left.  The individual T1-weighted image is then nonrigidly registered to the anatomical template.  Optionally, one can register the T1 template to the MNI152 template which resides in the same space as the FMRIB58 template.  Thus, the composite transform, carefully constructed without any FA-to-FA image registration, can take an individual subject's FA map (or other DTI-derived measure) to the desired standardized coordinate system.    }
\label{fig:workflow}
\end{center}
\end{figure*}

We first briefly describe the SyGN algorithm used to create anatomical templates.  
We then explain the the step-by-step process for aligning subject FA images to 
the anatomical template (which is facilitated with the various tools available 
in ANTs).  After describing the real image data used in this work, we show the 
issues raised affect common analysis scenarios.

%
%
%
%It was precisely due to these concerns that the TBSS framework was developed constituting an entire processing pipeline spanning from the reconstruction of the DTI tensorial data
%%to the calculation of the corresponding DTI-derived scalar measures (typically fractional anisotropy (FA)), to the nonrigid alignment of the population to a standard coordinate system and skeletonization, and, finally, 
%through to the statistical analysis for discerning differences across populations.  Major intermediate steps include the calculation of the corresponding DTI-derived scalar measures (typically fractional anisotropy (FA) \citep{Basser1996}), nonrigid alignment of the population to a standard coordinate system, and subsequent ``skeletonization'' of the mean of the aligned FA maps.  The alignment to a standard space is performed in such a way to remove the smoothing step associated with SPM.  Skeletonization and its projection across all registered DTI-derived scalar images is meant to accommodate the possibility of imperfectly aligned data.  However, as shown in a recent study \citep{Zalesky2011}, although the alignment error is reduced with the skeleton projection step, misalignments still remain post-projection.  Added to this is the potential confounding effect of minimizing data differences via nonrigid registration which are potentially the precise differences to be statistically quantified.

%Because spatial normalization is critical to obtaining valid results, we propose a modification of the standard TBSS pipeline directed at some of these residual concerns which we refer to as ANTs-TBSS.  Instead of DTI-derived scalar alignment via the images themselves (either random subject selection or locating an approximation to the ``mean subject'') or to a standard template (e.g. FMRIB58\_FA\_1mm.nii.gz which we denote as ``FMRIB58'' in the remainder of the text), we evaluate the substitution of an anatomical, i.e. T1-weighted, unbiased shape/intensity template constructed from the population images.  DTI-derived images can then be mapped to their corresponding T1-weighted image in a highly constrained manner and subsequently mapped to the template via nonrigid registration of the T1-weighted image.  This eliminates any possible complications associated with direct FA-to-FA image registration.

>>>>

\newcommand{\lstsetcpp}
{
\lstset{frame = tb,
        framerule = 0.25pt,
        float,
        fontadjust,
        backgroundcolor={\color{listlightgray}},
        basicstyle = {\ttfamily\scriptsize},
        keywordstyle = {\ttfamily\color{listkeyword}\textbf},
        identifierstyle = {\ttfamily},
        commentstyle = {\ttfamily\color{listcomment}\textit},
        stringstyle = {\ttfamily},
        showstringspaces = false,
        showtabs = false,
        numbers = none,
        numbersep = 6pt,
        numberstyle={\ttfamily\color{listnumbers}},
        tabsize = 2,
        language=[ANSI]C++,
        floatplacement=!h,
        caption={\small {\ttfamily CreateDTICohort} which implements the a ground truth DTI simulator \cite{van-hecke2009}.  The short command line menu which is invoked using the  `{\ttfamily {-}h}'  option.  The long command line menu is obtained by typing `{\ttfamily {-}{-}help}'},
        captionpos=b,
        }
}

@@

\title{Circularity in Voxel-Based Analysis}

@@

\begin{abstract}
Recent discussions within the neuroimaging community have highlighted the
problematic presence of selection bias in experimental design.  Although 
initially centering around the selection of voxels during the course of
fMRI studies, we demonstrate how this concept of circularity bias can 
potentially corrupt voxel-based analyses with a particular emphasis on
population studies of fractional anisotropy (FA).  
For such studies, template-based registration plays a critical role
in which a representative template serves as the normalized space for 
group alignment.  Following 
computation of FA, each subject's FA map is registered to the FA template prior to 
performing statistical comparisons between different groups within a population.
We illustrate how this analysis strategy can induce selection
bias which varies in strength with the similarity metric used in
registration.  Specifically, we show that the popular sum of squared
difference intensity metric  maximizes effect size, rather than
anatomical alignment, and consistently overestimates statistical
significance.  Consequently, we advocate a more cautious approach
that normalizes a population of FA images through the associated 
anatomical images.  One possibility, the Symmetric Group Normalization
strategy for unbiased template construction from anatomical (e.g.
T1-weighted) images, provides a valid and unbiased coordinate system for
statistical inference of FA differences.  
We establish this
outcome by using simulated data derived from normal controls from the International
Neuroimaging Data-sharing Initiative as well as a cohort of 16 traumatic
brain injury survivors compared with 17 control subjects.  
Alignment
comparisons with the normalization component of the popular tract-based
spatial statistics (TBSS) show that our anatomically-based template
approach produces good alignments despite the absence of any direct
FA-to-FA registration.  
%Of practical consideration we note that the
%software tools discussed in this work are provided to the public as part
%of the open source Advanced Normalization Tools (ANTs) repository.  
In conclusion, we recommend against circular analysis strategies that 
select image similarity metrics that are tied directly to the statistical 
functions used to assess population effects. 
\end{abstract}

\begin{keyword}
Advanced Normalization Tools (ANTs) \sep circularity\ diffusion tensor imaging \sep fractional anisotropy \sep template construction \sep voxel-based analysis
%% keywords here, in the form: keyword \sep keyword
\end{keyword}

@@

\section{Introduction}
Motivating the findings of a recent paper, the authors related their
puzzlement at the unusually high number of papers reporting 
significant correlations in fMRI studies \cite{Vul2009}. During the 
resulting investigation, the authors discovered the permeating
presence of a particular bias characterized by such related terms as ``circularity''
and ``non-independence''.
While related literature has focused on the presence of such bias in 
fMRI studies \cite{Vul2009,Vul2010} and, more generally, in neuroscience research 
\cite{Kriegeskorte2009}, we show how this can potentially extend to what has
become a fundamental neuroimaging analysis technique---voxel-based analysis.

Similarly, we were perplexed by the disparity in statistical results of one of 
our recent analyses \cite{Stone2011} using the same DTI data and same 
voxel-based analysis technique but different normalization approaches.  Despite 
distinctively better alignments with one approach, there was little 
difference in the statistical results between the two approaches.  
It was eventually discovered that the 
incommensurability between alignment quality and effect size in our
analysis was also a due to a subtle, yet significant circularity bias
which we explain below in the context of FA population studies.  
%Due to the public availability and widespread use of
%these voxel-based analysis tools we show 
%how this source of bias can enter into voxel-based studies focusing on
%population-based FA .

%Whereas the authors
%of \cite{Vul2009} were able to estimate reliability measures of
%both personality tests and fMRI and, thus, provide an upper bound 
%for possible correlations, the nature of the the analysis provided
%no indication as to the source of the disparity.

%State the objectives of the work and provide an adequate background, avoiding a detailed literature survey or a summary of the results.

The seminal work of Basser et al. \citep{Basser1994a,Basser1994} established diffusion tensor imaging (DTI) as a viable investigatory MRI technique.  DTI's sensitivity to brain architecture \citep{Basser1996,Assaf2008} enables promising analysis possibilities assessing neuro-structural differences in cross-population studies \citep{Kubicki2005,Arnone2008,Kantarci2010,Rametti2010}.
% including subjects with traumatic brain injury (TBI) \citep{Caeyenberghs2010,Jiang2010,Warner2010}.  
Two popular comprehensive software packages for assessing population FA differences include the SPM Matlab-based toolkit%
\footnote{
http://www.fil.ion.ucl.ac.uk/spm/
}
for voxel-based morphometry (VBM)
and the tract-based spatial statistics (TBSS) framework \citep{Smith2006}.%
\footnote{
http://www.fmrib.ox.ac.uk/fsl/tbss/index.html
} 
VBM \citep{Ashburner2001} continues to find application to FA studies \citep{Kakeda2010,Takao2010,Preziosa2011} despite concerns about its general validity \citep{Bookstein2001,Davatzikos2004} and with respect to DTI \citep{Jones2005,Chung2008}.  The TBSS framework was partially developed in response to these concerns in which voxel values are projected onto a template white matter skeleton for increased statistical power.  
%(e.g. \citep{Arnone2008,Bodini2009}).

Although there are substantive differences in normalization
between these and other analysis approaches, a core
commonality includes direct FA-to-FA registration using various similarity metrics.  The sum of squared differences (SSD) image similarity metric is perhaps the easiest to interpret (it drives the image intensity difference to zero), is computationally efficient (just compute the image difference and gradient), and is therefore widely used.  Some of the most popular registration methods (e.g. SPM, 
Demons \cite{Thirion1998}, FSL's FNIRT, AIR \citep{Woods1998}, and LDDMM \citep{Beg2005}) rely on this metric and yield reasonable performance levels \citep{Klein2009}.%
%\footnote{
%In addition to TBSS and SPM, other popular packages in the neuroimaging community 
%which use the SSD metric include AIR \citep{Woods1998} and LDDMM \citep{Beg2005}.
%}

However, suppose the SSD metric is used to find a set of $M$ transformations that map a population of $M$ FA images, $\{I_1, I_2, \ldots, I_M\}$, to a representative FA template, $J$, for statistical testing between groups. This results in the transformations,
$\boldsymbol{\mathcal{T}} = \{\mathcal{T}_1, \mathcal{T}_2,\ldots,\mathcal{T}_M \}$, which minimize the SSD metric over the population, i.e.
\begin{align}
  \boldsymbol{\mathcal{T}} = \underset{\mathcal{T}_1, \ldots, \mathcal{T}_M}{\operatorname{argmin}}
    \sum_{m=1}^M\sum_{n=1}^N \left( I^n_m( \mathcal{T}_m(\mathbf{x}) ) - J \right)^2.
\end{align}
Switching the order of the summations (since the transformations are mutually independent) and recognizing that a principal criterion for the selection of $J$ is such that it be a good approximation of the mean of the aligned images, it becomes apparent that the transformation solution
\begin{align}\label{eq:variance}
  \boldsymbol{\mathcal{T}} = \underset{\mathcal{T}_1, \ldots, \mathcal{T}_M}{\operatorname{argmin}}
    \sum_{n=1}^N \underbrace{\sum_{m=1}^M \left( I^n_m( \mathcal{T}_m(\mathbf{x}) ) - J \right)^2}_{
    \propto\mathrm{\,voxelwise\,variance}}
\end{align}
is that which minimizes the voxelwise group variance, or at least an approximation thereof.  Reducing the group variance will also directly affect the value of a standard population statistic, the Student's $t$-test.  Other standard similarity metrics will also exhibit this
trend to some degree as they minimize some local measure of intensity-based distance between the subject and the template.

Thus, instead of normalization based on anatomical alignment followed by
statistical testing, direct normalization of the images to be statistically compared 
explicitly conflates
anatomical alignment with optimizing the very statistical testing
results that one will ultimately use to assess hypotheses.  This immediately evokes a sense of circularity in the analysis \citep{Kriegeskorte2010}.
According to Kriegeskrote et al., ``An analysis is circular (or nonindependent) if it is based on data that were selected for showing the effect of interest or a related effect.''  As illustrated with
Eqn. \ref{eq:variance}, direct FA-to-FA template registration will produce transformations that align (or select) voxels such that the statistical testing result (or the effect of interest) is increased (assuming one is using Student's 
$t$-test).  While the SSD metric is able to provide visually attractive alignments across subjects, the voxel-level details (that is, which voxels are matched where) are strikingly important in terms of the effect on detection power.

Although various processing choices mitigate the effects of such bias,
such as Gaussian smoothing, nonparametric testing \citep{Rorden2007}, 
statistical analysis based on orthogonal projections onto the white matter skeleton 
\cite{Smith2006}, multiple comparisons corrections \cite{Nichols2003}, and the use of more robust similarity metrics, 
such choices are often made ad hoc and/or post hoc and, to our 
knowledge, not with the realization of the potential
for circularity bias described by Eqn. (\ref{eq:variance}).

Instead of direct FA-to-FA template mapping for 
alignment to the normalized space, we advocate a statistically 
independent alignment protocol in which the FA images are mapped to 
the common reference space via intra-subject FA-to-T1 mapping 
followed by T1-to-T1 template alignment thus eliminating all direct
FA-to-FA registrations. The key point is that the intensity value differences used 
to assess alignment (i.e. the similarity metric) in FA-to-T1 and T1-to-T1 
normalizations are statistically independent of the intensity value differences used for 
the post-alignment numerical analysis.  It should be noted that this discussion
is certainly not limited to studies involving FA.  From 
Eqn. (\ref{eq:variance}), it is evident that such bias can occur anytime the 
similarity metric (and the values employed) are non-independent of 
the statistical analysis.

Using simulated data, we first establish that this effect is real 
and warrants concern on the part of researchers using 
these techniques.  We then show how the general application
of a high performance \citep{Klein2009}, nonrigid registration algorithm 
\citep{Avants2011}, which forms the underlying machinery for an optimal 
template construction algorithm \citep{Avants2010}, can facilitate alignment 
between DTI-derived scalar images for cross-population studies.  
Although this type of alignment procedure is not uncommon 
(e.g. \cite{Lipton2008}), to our knowledge, this is the 
first time a connection has been made with the concept of 
circularity.

%The rest of the paper establishes the severity of 
%this bias in situations
% although scientific best 
%practices would dictate the reasonable removal of all 
%potential sources of bias (including circularity) regardless of 
%the general degree of severity.

\section{Methods and Materials}
% Provide sufficient detail to allow the work to be reproduced. 
% Methods already published should be indicated by a reference: 
% only relevant modifications should be described.

\subsection{Template Construction for Anatomically-Based Population Normalization}

\begin{figure*}
\begin{center}
\begin{tabular}{c}
  \includegraphics[width=170mm]{workflow.pdf}
%  \includegraphics[width=130mm]{workflow.pdf}
\end{tabular}
\caption{Possible ANTs-based TBSS workflow using SyGN.  In contrast to standard TBSS in which an individual subject's FA is registered to the FMRIB58 template
(transform represented with the blue arrow), our proposed modification first constructs an ANTs optimal anatomical template as illustrated in the center of the figure.  Each FA image is then rigidly mapped, with distortion correction, to the individual T1 using the white matter mask as shown on the left.  The individual T1-weighted image is then nonrigidly registered to the anatomical template.  Optionally, one can register the T1 template to the MNI152 template which resides in the same space as the FMRIB58 template.  Thus, the composite transform, carefully constructed without any FA-to-FA image registration, can take an individual subject's FA map (or other DTI-derived measure) to the desired standardized coordinate system.    }
\label{fig:workflow}
\end{center}
\end{figure*}

In general, normalization to a standard coordinate system to
reduces intersubject variability in population studies.  Various
approaches exist for determining the normalized space such as selection
of a pre-existing template based on a single subject, e.g. the Talairach
atlas \citep{Talairach1988}, or a publicly available averaged group of
subjects, e.g. the MNI \citep{Collins1994} or ICBM \citep{Mazziotta1995}
templates.  
%One challenge with standard templates is that they may inadvertantly bias one's results by enabling better normalization of subjects to which the template is more similar.  This issue is exacerbated when dealing with populations that have high variance (e.g. due to disease) and/or when one's normalization method is low-dimensional (not flexible enough to capture large shape differences). 
Population-specific templates (or optimal templates) alleviate some of
these issues by deriving a most representative image from the population
\citep{Good2001}.  Large deformation registration algorithms also reduce
this confound by being less sensitive to the deformation distance
between subject and target.  Some approaches combine both advantages,
for instance, the diffeomorphic approach of Joshi et al. employs the
SSD metric and a shape distance to bring the subject group of images
into alignment \citep{Joshi2004}.  Variants
include extension to multiple modalities \citep{Lorenzen2006} and small deformations
\citep{Geng2009}.  These approaches iteratively minimize group difference in ``congealing''
towards a representative image template \citep{Learned-Miller2006}.
In contrast, Avants et al. explicitly model the geometric component of the 
normalized space during optimization.  Coupling the intrinsic symmetry of 
SyN pairwise registration \citep{Avants2011} and an
optimized sharpening/averaging of the template appearance, Symmetric Group
Normalization (SyGN) is a powerful framework for producing optimal population-specific
templates \citep{Avants2010} with arbitrary similarity metric choice.

%Given a set of representative images, 
%$\{I_1, I_2, \ldots, I_M\}$, optimization involves finding the set of paired
%diffeomorphic transformations, $\left\{\left(\phi^1_1, \phi^1_2\right), 
%\left(\phi^2_1, \phi^2_2\right), \ldots, \left(\phi^M_1, \phi^M_2\right) \right\}$,
%the optimal template appearance, $J^*$, and corresponding coordinate system, $\psi(\mathbf{x})$,
%which minimize the following cost function:
%\begin{align}
%  \sum_{m=1}^M \left[
%    D\left( \psi(\mathbf{x}), \phi^m_1(\mathbf{x}, 1) \right) + 
%    \Pi \left( I_m\left(\phi^m_2(\mathbf{x}, 0.5) \right), J^*\left(\phi^m_1(\mathbf{x}, 0.5) \right) \right)
%    \right]
%\end{align}
%where $D$ is the diffeomorphic shape distance, 
%\begin{align}
%  D\left(\phi(\mathbf{x}, 0), \phi(\mathbf{x}, 1)\right) = \int_{0}^1 \| v(\mathbf{x}, t) \|_L dt, 
%\end{align}
%dependent upon the choice of the linear operator, $L$, and
%$v$ is the diffeomorphism-generating velocity field, 
%\begin{align}
%  v\left(\phi(\mathbf{x}, t), t \right) = \frac{d\phi(\mathbf{x}, t)}{dt},\,\,\, \phi(\mathbf{x}, 0) = \mathbf{x}.
%\end{align}
%$\Pi$ is the choice of similarity metric, often cross-correlation \citep{Avants2008a}, calculated in the 
%virtual domain midway between each individual image and the current estimate of the template. 
%
%With initial assignment of $\left\{\left(\phi^m_1, \phi^m_2\right)\right\}$ and $\psi(\mathbf{x})$ 
%to identity, iterative optimization
%involves estimating the pairwise transformations, estimation of the optimal template appearance, and 
%updating $\psi(\mathbf{x})$ by averaging the current estimate of $\left\{\phi_1^m\right\}$.  Additional details are given in \cite{Avants2010}.  
%For the interested reader, a set of scripts has been created and is distributed with the ANTs library for facilitating template creation.

Our proposed workflow applied to anatomically-based population normalization is illustrated in Fig. \ref{fig:workflow}.  Instead of direct FA-to-FA registration, our ANTs-based modification can be described as follows:
